Local Type Inference for Polarised System F with Existentials

### 2023-10-17 oopsla24/tex/02-overview.tex
Wip: declarative system
wip

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similarly to the \CBPV system \cite{levy2006:cbpv}.
The negative types represent computations, and the positive types represent values:
\begin{itemize}
\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables 
  $[[pas]]$. The polarities are chosen to follow the definition of functions.
\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term-level corresponds to a pure computation $[[return v]]$.
\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
  binds negative variables in a positive type $[[iP]]$. 
\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
  thunks a computation, which at the term-level corresponds to $[[ {c} ]]$.
\end{itemize}

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }
    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

\paragraph{Definitional Equalities}
For simplicity, we assume alpha-equivalent terms equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

@@

In \cref{fig:declarative-terms}, we define the the language of terms of 
\fexists. The language combines \systemf with the \CBPV approach.

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
        \ilyam{why no negatives?}
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

>>>>

The types of \fexists are given in \cref{fig:declarative-types}.
They are stratified into two syntactic 
categories (polarities): positive and negative,  
similarly to the \CBPV system \cite{levy2006:cbpv}.
The negative types represent computations, and the positive types represent values:
\begin{itemize}
\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables 
  $[[pas]]$. The polarities are chosen to follow the definition of functions.
\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$.
\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
  binds negative variables in a positive type $[[iP]]$. 
\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
\end{itemize}

\begin{figure}[h]
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottiN\ottinterrule
    }\\
    \ottgrammartabular{
      \ottiP\ottinterrule
    }
    \columnbreak
  \end{multicols}
  \caption{Declarative Types of \fexists}
  \label{fig:declarative-types}
\end{figure}

\paragraph{Definitional Equalities}
For simplicity, we assume that alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we equate
$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
and lift these equations transitively and congruently 
to the whole system.

@@

In \cref{fig:declarative-terms}, we define the language of terms of 
\fexists. The language combines \systemf with the \CBPV approach.

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
        \ilyam{why no negatives? Following CBPV}
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ operate similarly to 
        the bind of a monad: they take a suspended computation $[[v]]$,
        apply it to a list of arguments, bind the result 
        (which is expected to be pure) to a variable $[[x]]$,
        and continue with a computation $[[c]]$.
        If the resulting type of the application is unique, 
        one can omit the type annotation, as in the second form:
        it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

### 2023-10-17 oopsla24/tex/03-declarative-system.tex
wip

@@

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
the subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of the subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
The instantiation is modeled by substitution $[[σ]]$.
The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.
This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftDLabel} can be relaxed to the covariant form,
 as we will discuss in \cref{todo}.

@@

A property that we extensively use is that the subtyping is reflexive and transitive. 
Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.

\begin{property}[Subtyping is well-defined]
  The subtyping forms a preorder and is preserved under substitutions. 
\end{property}

\subsection{Equivalence and Normalization}

The subtyping relation forms a preorder on types,
and thus, it induces an equivalence relation \aka \emph{bicoercibility}.
The declarative specification of subtyping must be defined up to this equivalence.
Moreover, the algorithms we use must withstand changes in input types within the equivalence class.
To deal with this, we use normalization---a function that uniformly selects a representative of the equivalence class.

Using normalization gives us two benefits:
\begin{enumerate*}
  \item [(i)] we do not need to modify significantly standard operations such as unification to withstand non-trivial equivalence, and
  \item [(ii)] once the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
    while the rest of the algorithm remains the same.
\end{enumerate*}

@@

  \begin{multicols}{2}
  \ottdefnNrmNNorm{}
  \columnbreak\\
  \ottdefnNrmPNorm{}
  \end{multicols}
  \caption{Type Normalization}
  \label{fig:type-nf}
\end{figure}

@@

The declarative type inference system is shown in \cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to have the type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.
To infer the resulting type of an application, the
applicative let rules refer to the \emph{Application typing} judgment.
It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which is read as ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to have the type $[[iM]]$''.

\begin{figure}[h]
  \ottdefnDTNInf{}

  \hfill

  \begin{multicols}{2}
  \ottdefnDTPInf{}
  \\
  \ottdefnDTSpinInf{}
  \end{multicols}
  \hfill

@@

\paragraph{Non-bidirectionality}
  Notice that all the rules in our system are inferring.

\paragraph{Typing up to equivalence}
  As discussed in \cref{todo}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  The application typing is defined up to equivalence as well,
  although it has no special rules for that.

@@

\paragraph{Application of a polymorphic type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  taken out of thin air. The algorithm we present in \cref{todo} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

@@

\subsection{Properties of the Declarative System}

Now we present selected properties of the declarative system,
which are important for the correctness of the algorithm.

As we discussed in \cref{todo}, the equivalence induced by the subtyping is 
not trivial. To deal with it without modifying standard operations such as unification, 
we use normalization. This way, it is important for the normalization to be 
sound and complete \wrt the equivalence:

\begin{property}[Correctness of normalization]
  For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$.
\end{property}

However, the greatest lower bound of two positive types is not always defined.
Let us consider a positive type $[[iP]]$,
and two different supertypes $[[iP1]]$ and $[[iP2]]$
(\ie  $[[· ⊢ iP1 ≥ iP]]$, $[[· ⊢ iP2 ≥ iP]]$, and $[[nf(iP1) ≠ nf(iP2)]]$.
Let us also consider an arbitrary negative type $[[iN]]$.

>>>>

\label{sec:declarative-system}

@@

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
the subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of the subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively.

The instantiation of quantifiers is modeled by substitution $[[σ]]$.
The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.
This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
 as we will discuss in \cref{todo}.

@@

\subsubsection{Properties of the Declarative Subtyping}
A property that we extensively use is that the subtyping is reflexive and transitive,
and agrees with substitution.

\begin{property}[Subtyping forms a preorder]
  Let us say that two types $[[iN1]]$ and $[[iN2]]$ are in the subtyping relation if there 
  exists a context $[[Γ]]$ such that $[[Γ ⊢ iN1 ≤ iN2]]$;
  symmetrically, two types $[[iP1]]$ and $[[iP2]]$ are in the subtyping relation
  if there exists $[[Γ]]$ such that $[[G ⊢ iP2 ≥ iP1]]$.
  Then the subtyping relation defined this way is reflexive and transitive.
\end{property}

\begin{property}[Subtyping agrees with substitution]
  Suppose that  $[[σ]]$ is a substitution such that $[[Γ2 ⊢ σ : Γ1]]$. 
  Then
  \begin{enumerate}
    \item [$-$] $[[Γ1 ⊢ iN ≤ iM]]$ implies $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$, and
    \item [$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ implies $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$.
  \end{enumerate}
\end{property}

Moreover, any two \emph{positive} types have the least upper bound, which makes
the positive subtyping semilattice.
The positive least upper bound can be found algorithmically,
which we will discuss in the next section.

\begin{property}[Positive Least Upper Bound exists]
  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
  well-formed in $[[Γ]]$.
  Then there exists the least common supertype--- a type $[[iP]]$ such that
  \begin{itemize}
    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
      $[[Γ ⊢ iQ ≥ iP]]$.
  \end{itemize}
\end{property}

\paragraph{Negative Greatest Lower Bound does not exist}
However, the symmetric construction---the greatest lower bound of two negative types
does not always exist. Let us consider the following counterexample.
Let us consider the following types: 
\begin{itemize}
  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
    none of the types is equivalent to $[[iQ]]$.
\end{itemize}
What is the greatest common subtype of  
$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
One of the common subtypes is $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$,
which, however is not the greatest one.

One can find two greater candidates:
$[[iM1 = ∀α⁺,β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and $[[iM2 = ∀α⁺,β⁺. β⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$.
Instantiating $[[α⁺]]$ and $[[β⁺]]$ with $[[iQ]]$ ensures 
that both of these types are subtypes of $[[iQ → ↓↑iQ → ↓↑iQ → iN]]$;
instantiating $[[α⁺]]$ with $[[iP1]]$ and $[[β⁺]]$ with $[[iP2]]$
demonstrates the subtyping with $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$,
as $[[iP]]$ is a subtype of both $[[iP1]]$ and $[[iP2]]$.

By analyzing the inference rules, we can prove that
both $[[iM1]]$ and $[[iM2]]$ are maximal common subtypes.
Since $[[iM1]]$ and $[[iM2]]$ are not equivalent,
it means that none of them is the greatest.

\subsection{Equivalence and Normalization}
\label{sec:decl-equivalence}

The subtyping relation forms a preorder on types,
and thus, it induces an equivalence relation \aka bicoercibility 
\cite{tiuryn1995:bicoercibility}.
The declarative specification of subtyping must be defined up to this equivalence.
Moreover, the algorithms we use must withstand changes in input types within the equivalence class.
To deal with non-trivial equivalence, 
we use normalization---a function that uniformly selects a representative of the equivalence class.

Using normalization gives us two benefits:
\begin{enumerate*}
  \item [(i)] we do not need to modify significantly standard operations such as unification to withstand non-trivial equivalence, and
  \item [(ii)] if the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
    while the rest of the algorithm remains the same.
\end{enumerate*}

In our system, equivalence is reacher than equality. Specifically,
\begin{enumerate}
  \item[(ii)] one can introduce redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
  but not equal to $[[∀α⁺.↑α⁺]]$;
  \item[(i)] one can reorder quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
  $[[∀α⁺,β⁺.β⁺ → α⁺ → γ⁻]]$; 
  \item[(iii)] the transformations (i) and (ii) can happen at any
    position in the type.
\end{enumerate}

It turns out that the transformations (i-iii) are complete, 
in the sense that they generate the whole equivalence class.
This way, to normalize the type, one must
\begin{enumerate}
  \item [(i)] remove the redundant quantifiers,
  \item [(ii)] reorder the quantifiers to the canonical order,
  \item [(iii)] do the procedures (i) and (ii) recursively on the subterms.
\end{enumerate}

The normalization algorithm is shown in \cref{fig:type-nf}.
The steps (i-ii) are implemented by the ordering function, which takes a set of variables $[[varset]]$
and a type and returns a list of variables from $[[varset]]$ that occur in the type in the order of their
first occurrence. Its formal definition can be found in \cref{todo}.

@@

  \begin{multicols}{2}
  \ottdefnNrmNNorm{}
  \\
  $[[ord varset in iN]]$ returns a list of variables
  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
  \columnbreak\\
  \ottdefnNrmPNorm{}
  \\
  $[[ord varset in iP]]$ returns a list of variables
  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
  \end{multicols}

  \caption{Type Normalization Procedure} 
  \label{fig:type-nf}
\end{figure}

For the normalization procedure, 
we prove soundness and completeness \wrt the equivalence relation.
\begin{property}[Correctness of normalization]
  \hfill \\
  \begin{itemize}
    \item[$-$] For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$;
    \item[$+$] analogously, for $[[iP]]$ and $[[iQ]]$ well-formed in $[[Γ]]$,
  $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
  \end{itemize}
\end{property}

@@

The declarative specification of the type inference is shown in 
\cref{fig:declarative-inference}.
The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the term $[[v]]$ is allowed to have the type $[[iP]]$'',
where $[[Φ]]$---the variable context---is defined standardly as
a set of pairs of the form $[[x : iP]]$. 
The negative typing judgment is read similarly.

The \emph{Application typing} judgment
infers the type of the application of a function to a list of arguments.
It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
is allowed to have the type $[[iM]]$''.

\begin{figure}[h]

  \begin{multicols}{2}
  \ottdefnDTNInf{}
  \columnbreak\\
  \ottdefnDTPInf{}
  \ottdefnDTSpinInf{}
  \end{multicols}
  \hfill

@@

Let us discuss the rules of the declarative system in more detail.

\paragraph{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows to infer
  the type of a variable from the context. 
  In literature can be found another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

\paragraph{Return and Thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  add the corresponding shifts to the type of the body.

@@

\paragraph{Typing up to Equivalence}
  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
  induces a non-trivial equivalence relation on types. 
  The system must not distinguish between equivalent types,
  and thus, type inference must be defined up to equivalence. 
  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
  and \ruleref{\ottdruleDTNEquivLabel}.
  They allow one to replace the inferred type with an equivalent one.

\paragraph{Application to an Empty List of Arguments}
  The base case of the application type inference is 
  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
  If the head of the type $[[iN]]$ is applied to no arguments, 
  the type of the result is allowed to be $[[iN]]$ or any 
  equivalent type. We need to relax this rule up to equivalence
  to ensure the corresponding property globally:
  the inferred application type can be replaced with an equivalent one.
  Alternatively, we could have added a separate rule similar to 
  \ruleref{\ottdruleDTPEquivLabel}, however, 
  the local relaxation is sufficient to prove the global property.

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  The complexity of the system is hidden in the rules, 
  whose output type is not immediately defined by their input
  and the output of their premises 
  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
  the instantiation of the quantifying variables is
  not known a priori. The algorithm we present in \cref{todo} 
  delays this instantiation until more information about it 
  (in particular, typing constraints) is collected.

  To ensure the priority of application between this rule and 
  \ruleref{\ottdruleDTEmptyAppLabel}, we also check that 
  the list of arguments is not empty.

@@

\subsubsection{Declarative Typing Properties}
  An important property that the declarative system has is
  that the declarative specification is correctly defined for
  equivalence classes.

\begin{property}[Declarative Typing is Defined up to Equivalence]
  Let us assume that $[[Γ ⊢ Φ1 ≈ Φ2]]$, \ie  
  the corresponding types assigned by $[[Φ1]]$ and
  $[[Φ2]]$ are equivalent in $[[Γ]]$.
  Also, let us assume that 
  $[[Γ ⊢ iN1 ≈ iN2]]$, $[[Γ ⊢ iP1 ≈ iP2]]$,
  and $[[Γ ⊢ iM1 ≈ iM2]]$. Then
  \begin{itemize}
    \item [$-$] $[[Γ ; Φ1 ⊢ c : iN1]]$ holds if and only if $[[Γ ; Φ2 ⊢ c : iN2]]$,
    \item [$+$] $[[Γ ; Φ1 ⊢ v : iP1]]$ holds if and only if $[[Γ ; Φ2 ⊢ v : iP2]]$, and
    \item [$\bullet$] $[[Γ; Φ1 ⊢ iN1 ● args ⇒> iM1]]$ holds if and only if $[[Γ; Φ2 ⊢ iN2 ● args ⇒> iM2]]$.
  \end{itemize}
\end{property}

\ilyam{Other properties?}
